A Triple Origin for Twin Blue Stragglers in Compact Binaries

### 2018-12-21 BSS7782.tex
Updated paper

We propose a formation channel for Binary 7782, and compact double BS
binaries in general, which involves mass transfer from an outer
tertiary companion on to an inner (ms, ms) binary.  In
section~\ref{sect:dyn}, we constrain the range of initial (i.e., pre-mass
transfer) orbital parameters for a hypothetical outer tertiary
companion, using a combination of dynamical and stellar
evolution-based constraints.  In Section~\ref{sims} we present the
numerical simulations used to study the mass transfer process in our
hypothetical triple system, computed using the
\texttt{AMUSE}\cite{AMUSE} software package, in order to study the
evolution of the inner and outer orbital parameters during mass
transfer.  We summarize and discuss the implications of our results
for compact double BS binaries and, more generally, mass transfer in
stellar triples in Section~\ref{discussion}.

@@

\subsection{AMUSE} \label{amuse}

We perform simulations of a triple star system for which the outer
star overfills its Roche lobe while the inner binary remains
detached. The calculations start by evolving the three stars to the
same age, which is selected such that the outer-most star fills its
Roche lobe.  First order constraints for the initial conditions are
derived in the previous \S\, but in \S\,\ref{Sect:ICs} we summarize
the adopted initial conditions for the more detailed simulations.

Roche-lobe overflow in triples is modeled using a coupled integrator
to follow the complex hydrodynamics of the mass transfer from the
Roche-lobe filling outer star to the inner binary, while keeping track
of the gravitational dynamics of the stars.  The equations of motion
of the inner binary are solved using the symplectic direct N-body
integrator \texttt{Huayno} \citep{2012NewA...17..711P}. The
hydrodynamics is performed with the smoothed-particles hydrodynamics
code \texttt{Gadget2} \citep{2000ascl.soft03001S}, using an adiabatic
equation of state.  The two inner binary stars are treated as point
masses, but we allow them to accrete mass and angular momentum from
the gas liberated by the outer star.  This is realized using
sink-particles that co-move with the mass points in the gravity
code. While the inner two stars accrete mass, they also accrete the
corresponding amount of angular momentum from the gas.  The the N-body
integrator accounts of this appropriately.  For the radius of the sink
particles we adopt $2 R_\odot$ for both stars.

@@

\subsection{Initial Conditions} \label{Sect:ICs}

We adopt initial component masses of $m_{\rm 1} = 1.1$ M$_{\rm \odot}$
and $m_{\rm 2} = 0.9$ M$_{\rm \odot}$ for the inner binary components,
and between $m_{\rm 3} = 1.2$ and $m_{\rm 3} = 1.4$ M$_{\rm \odot}$
for the tertiary star.  We evolve the tertiary star using the MESA
stellar-evolution code \cite{2011ApJS..192....3P} to a radius of about
1\,au.  By this time the star has lost 0.13\,\MSun\, via a stellar
wind.  We subsequently turn the stellar-evolution model, structure,
temperature and composition, into a smoothed-particles representation
using the module {\tt StellarModelInSPH} in AMUSE (see chapter 4 in
\cite{AMUSE}).  We replace the stellar core by a point mass to prevent
that all the resolution will be in the central portion of the star
instead of the envelope. In a following step we relax the star using
the hydrodynamics solver. This relaxation process is realized in 100
steps during which we reduce the velocity dispersion of individual SPH
particles to a glasseas structure \citep[see, for example, \S\,3.3 on
  page 40 in][]{1994astro.ph.10043W}. During this procedure the
gaseous envelope of the star tends to expand by about 20\%.  To
detemrine the radius of the eventual star we calculate Lagrangian
radii and use the distance to the stellar center which contains 90\%
of it's mass. This 90\% mass-radius we consider as the stellar radius,
and match it with the Roche-lobe of the outer orbit.

With these parameters the orbital separation of the outer binary
becomes $\sim 3.86$\,au, and we adopt the outer orbit to be circular.
We choose a relative initial inclination between the inner and the
outer orbit of $i = 0^\circ$, $i = 9^\circ$ and $i = 180^\circ$.  The
first two choices are motivated by observed stellar triples composed
of low-mass stars which have been observed to favor co-planar
configurations \citep{moe18}. To test the effect of a retrogade
orbiting triple we include the $i = 180^\circ$ case.
% NL: What are the initial orbital separations?  Why this choice of
% initial inclination? - I added a sentence and reference to justify
% it.
In fig.\,\ref{fig:topview_at_t0} we present a top view of the triple
system at the onset of Roche-lobe overflow.

@@

@@

\section{Summary and Discussion} \label{discussion}

@@

\item For the inner binary, the rotational axes of the BSs should be aligned with each other and the orbital plane of the outer tertiary WD.  This is because accretion onto the BS progenitors proceeds via an accretion disk, that forms at the circularization radius and that has an orbital plane aligned with that of the outer tertiary.

@@

\item Twin BSs in compact binaries formed from the proposed mechanism should be more frequent in younger clusters with ages $\lesssim$ 4-6 Gyr.  This is because clusters with a main-sequence turn-off mass $\lesssim$ 1.2 M$_{\odot}$ have convective envelopes \citep[e.g.][]{iben91,maedoer09}, and a radiative envelope for the donor in a mass transferring binary ensures stable accretion on to the accretor.

@@

Finally, in future work, we intend to explore the implications of the mechanism proposed here for other types of stellar triples.  Our results highlight that specific theoretical predictions can often be made.  For example, if two white dwarfs form the inner compact binary, than circumbinary accretion could push one of them above the Chandrasekhar mass limit.  Upon explosion, its WD companion would be ejected at roughly the orbital velocity at the time of explosion, producing a hypervelocity WD.  This predicts hot hypervelocity WDs with masses very close to the Chandrasekhar mass limit for this scenario.  Similarly, if the two WDs are replaced with two neutron stars (NSs), then one NS could accrete to surpass the critical supernova limit, upon which time it explodes to become a black hole and receives a strong natal kick.  The NS companion would have a very comparable mass to the newly formed black hole progenitor at the time of explosion, and could potentially be used to constrain the maximum NS mass.  More work needs to be done here to evaluate whether or not there is any parameter space for the products of this scenario that will allow for a mass determination for the original NS companion in the putative compact NS-NS binary.

@@

N.W.C.L. acknowledges support from a Kalbfleisch Fellowship at the
American Museum of Natural History.  SPZ would like to thank Norm
Murray and CITA for their hospitality during my long-term visit.

>>>>

We propose a formation channel for Binary 7782, and compact double BS
binaries in general, which involves mass transfer from an outer
tertiary companion on to an inner (ms, ms) binary.  In
section~\ref{sect:dyn}, we constrain the range of initial (i.e., pre-mass
transfer) orbital parameters for a hypothetical outer tertiary
companion, using a combination of dynamical and stellar
evolution-based constraints.  In Section~\ref{sims} we present the
numerical simulations used to study the mass transfer process in our
hypothetical triple system, computed using the
\texttt{AMUSE}\cite{AMUSE} software package, in order to study the
evolution of the inner and outer orbital parameters during mass
transfer.  We summarize and discuss the implications of our results
for compact double BS binaries and, more generally, mass transfer in
stellar triples in Section~\ref{sect:discussion}.

@@

We perform simulations of a triple star system for which the outer
star overfills its Roche lobe while the inner binary remains
detached. The calculations start by evolving the three stars to the
same age, which is selected such that the outer-most star fills its
Roche lobe.  First order constraints for the initial conditions are
derived in the previous \S. In the following two sections we describe
how we setup these simulations and then discuss the results. The
calculations are performed using the

\subsection{Setting-up the simulations}

We adopt initial masses of $m_{\rm 1} = 1.1$ M$_{\rm \odot}$ and
$m_{\rm 2} = 0.9$ M$_{\rm \odot}$ for the inner binary components, and
between $m_{\rm 3} = 1.2$ and $m_{\rm 3} = 1.4$ M$_{\rm \odot}$ for
the tertiary star.  We evolve the tertiary star using the MESA
stellar-evolution code \cite{2011ApJS..192....3P} to a radius of about
1\,au, at which point we assume it to overfill it's Roche lobe.  By
this time the star has lost 0.13\,\MSun\, via a stellar wind.

The stellar-evolution model, its structure, temperature and
composition profiles are turned into a smoothed-particles
representation using the module {\tt StellarModelInSPH} in AMUSE (see
chapter 4 in \cite{AMUSE}).  We follow the same procedure as described
in \cite{2014MNRAS.438.1909D} for simulating the future of the triple
system \chi Tau (HD 97131) in which the outer-most star overfills it
Roche lobe to an inner binary.  After generating the hydrodynamical
representation of the donor star we replace the stellar core by a
point mass to prevent the majority of the resolution to be confined in
the star's central region.  In a following step we relax the star
using the hydrodynamics solver. This relaxation process is realized in
100 steps during which we reduce the velocity dispersion of individual
SPH particles to a glasseas structure \citep[see, for example, \S\,3.3
  on page 40 in][]{1994astro.ph.10043W}. During this procedure the
gaseous envelope of the star tends to expand by about 20\%.  To
detemrine the radius of the eventual star we calculate Lagrangian
radii and use the distance to the stellar center which contains 90\%
of it's mass. This 90\% mass-radius we consider as the stellar radius,
and match it with the Roche-lobe of the outer orbit.

With these parameters the orbital separation of the outer binary
becomes $\sim 3.86$\,au, and we adopt the outer orbit to be circular.
We choose a relative initial inclination between the inner and the
outer orbit of $i = 0^\circ$, $i = 9^\circ$, $i = 90^\circ$ and $i =
180^\circ$.  The first two choices are motivated by observed stellar
triples composed of low-mass stars which have been observed to favor
co-planar configurations \citep{moe18}. To test the effect of highly
inclined and retrogade orbits we include the $i = 90^\circ$ and $i =
180^\circ$ cases.  In fig.\,\ref{fig:topview_at_t0} we present a top
view of the triple system at the onset of Roche-lobe overflow.

Roche-lobe overflow in triples is modeled using a coupled integrator
to follow the complex hydrodynamics of the mass transfer from the
Roche-lobe filling outer star to the inner binary, while keeping track
of the gravitational dynamics of the stars.  The equations of motion
of the inner binary are solved using the symplectic direct N-body
integrator \texttt{Huayno} \citep{2012NewA...17..711P}. The
hydrodynamics is performed with the smoothed-particles hydrodynamics
code \texttt{Gadget2} \citep{2000ascl.soft03001S}, using an adiabatic
equation of state.  The two inner binary stars are treated as point
masses, but we allow them to accrete mass and angular momentum from
the gas liberated by the outer star.  This is realized using
sink-particles that co-move with the mass points in the gravity
code. While the inner two stars accrete mass, they also accrete the
corresponding amount of angular momentum from the gas (see chapter 5
in \cite{AMUSE}).  The the N-body integrator accounts of this
appropriately.  For the radius of the sink particles we adopt $10
R_\odot$ for both stars.

@@

@@

\cite{2018arXiv181208175M}

@@

with the accretion of mass both stars in the inner binary also accrete
angular momentum.  By the end of the simulation the spins of the two
blue stragglers is aliened along the orbital angular momeumt axis with
an angle of $90.0^\circ$ for the primary star and $93.4^\circ$ for the
secondary star with respect to the argument of pericenter of the inner
orbit.  By the end of the simulations the spin of the primary is about
50.5 per day, and 41.5 per day for the secondary star. But by this
time both stars have accreted only a small portion, and we expect that
the final spins of both stars is close to break-up.

\section{Discussion} \label{sect:discussion}

In order to study the T-tauri binaries V4046 Sgr and DQ Tau
\cite{2011MNRAS.413.2679D} perform a series of 2D hydrodynamical
simulations of circum binary disks.  They studied the two observed
T-tauri system V4046 Sgr and DQ~Tau.

For V4046 Sgr, for which the two stars have comparable masses as in
our calculation in an ciruclar orbit with a period of only 2.4 days
they find tha the inner binary accretates at a rate of $\sim
0.028$\,\MSun/Myr.  For DQ~Taur, which is composed of lower-mass stars
($m_1 = m_2 \seimq 0.55$\,\MSun) in an eccentric ($e\simeq 0.556$)
orbit of $\sim 15.8$\,days they find and accretion rate onto the inner
binary of $\sim 0.027$\,\MSun/Myr.  These values are in the same range
as in our calculations, which result in an accretion rate of the inner
binary of 0.027--0.058\,\MSun/Myr (on average measurd over a period of
about 3000 days).  The highest accretion rate we obtain for the
simulations with an eccentric inner orbit. Here we applied a small
correction to compensate for the larger accretion radii adopted for
the two stars of 10\,\RSun\, compared to 6\,\RSun\, in
\cite{2011MNRAS.413.2679D}. Interestingly, however,
\cite{2011MNRAS.413.2679D} find that the primary star in V4046 Sgr
accretes at an 8\% higher rate than the secondary star, whereas in our
case the secondary star accretes at a higher rate than the primary
star by about 1\% to 12\%.  Higher accretion rates in the secondary
star are realized for eccentric and retrogade inner orbits.

\section{Summary} \label{sect:conclusions}

@@

\item For the inner binary, the rotational axes of both the BSs should
  be aligned with each other and the orbital plane of the outer
  tertiary WD.  This is because accretion onto the BS progenitors
  proceeds via an accretion disk, that forms at the circularization
  radius and that has an orbital plane aligned with that of the outer
  tertiary.

@@

\item Twin BSs in compact binaries formed from the proposed mechanism
  should be more frequent in younger clusters with ages $\lesssim$ 4-6
  Gyr.  This is because clusters with a main-sequence turn-off mass
  $\lesssim$ 1.2 M$_{\odot}$ have convective envelopes
  \citep[e.g.][]{iben91,maedoer09}, and a radiative envelope for the
  donor in a mass transferring binary ensures stable accretion on to
  the accretor.  \simon{Can we quantify this?}

@@

\simon{Why would we try to sell already what we plan on doing. I'de prefer to keep this quet.}
Finally, in future work, we intend to explore the implications of the
mechanism proposed here for other types of stellar triples.  Our
results highlight that specific theoretical predictions can often be
made.  For example, if two white dwarfs form the inner compact binary,
than circumbinary accretion could push one of them above the
Chandrasekhar mass limit.  Upon explosion, its WD companion would be
ejected at roughly the orbital velocity at the time of explosion,
producing a hypervelocity WD.  This predicts hot hypervelocity WDs
with masses very close to the Chandrasekhar mass limit for this
scenario.  Similarly, if the two WDs are replaced with two neutron
stars (NSs), then one NS could accrete to surpass the critical
supernova limit, upon which time it explodes to become a black hole
and receives a strong natal kick.  The NS companion would have a very
comparable mass to the newly formed black hole progenitor at the time
of explosion, and could potentially be used to constrain the maximum
NS mass.  More work needs to be done here to evaluate whether or not
there is any parameter space for the products of this scenario that
will allow for a mass determination for the original NS companion in
the putative compact NS-NS binary.

@@

N.W.C.L. acknowledges support from a Kalbfleisch Fellowship at the
American Museum of Natural History.  SPZ would like to thank Norm
Murray and CITA for their hospitality during my long-term visit.  This
work was supported by the Netherlands Research School for Astronomy
(NOVA), NWO (grant \# 621.016.701 [LGM-II]).
